Binning Strategies and Related Loss for Large Data

### 2015-04-13 Paper.Rnw
Move/resize airline plots and small edits to their descriptions.

\begin{figure}[hbtp]\centering
\includegraphics[keepaspectratio=true, height=.43\textheight]{images/AirlineStdBinning5mins.pdf}
\includegraphics[keepaspectratio=true, height=.43\textheight]{images/AirlineStdBinning15mins.pdf}
\caption{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b. 15-minute bins produce an even more coarse summary of the data.}\label{airline5mins-binning}
\end{figure}
Binning data by five-minute intervals produces a more high-level summary of the relationship between actual and scheduled departure time, though it necessarily obscures some of the finer details. In addition, binning data by five minute intervals reduces the size of the data set to a much more manageable 19,787 observations  which can be easily manipulated on probably any modern computer. Binning by 15-minute intervals reduces the data set to a nearly-trivial 3,575 observations, but the graphical summary becomes granular and less appealing at that resolution.

@@

# binnedRdm <- qplot(data=air.binned2.rdm, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
# ggsave("AirlineRdmBinning15mins.png", binnedRdm, width=8, height=8, dpi=600)
@
\section{Conclusion}
Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system. The loss due to binning is the cost of reducing the data set to a more manageable size, but some of that loss can be recovered with additional computational investment.

We have presented two algorithms for binning; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient. The choice of algorithm is a trade-off between speed and accuracy, and may be highly data set dependent. It is possible that there are computational modifications which could improve the accuracy of the random algorithm, but these modifications are likely to require additional computational operations.

% \begin{subsection}{Numerical Comparison of Binning Algorithms}
<<PrintTable,echo=F,include=FALSE, dependson='LossTableGraphs'>>=
  print(tab, include.rownames=FALSE)
@
% \end{subsection}

>>>>

\begin{figure}[hbtp]\centering
\includegraphics[width=.49\linewidth,keepaspectratio=true]{images/AirlineStdBinning5mins.pdf}
\includegraphics[width=.49\linewidth,keepaspectratio=true]{images/AirlineStdBinning15mins.pdf}
\caption{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b. 15-minute bins produce an even more coarse summary of the data.}\label{airline5mins-binning}
\end{figure}

Figure~\ref{airline5mins-binning} displays the binned scatterplots for the flight departure data with larger bins. Binning data by five-minute intervals produces a more high-level summary of the relationship between actual and scheduled departure time, though it necessarily obscures some of the finer details. In addition, binning data by five minute intervals reduces the size of the data set to a much more manageable 19,787 observations  which can be easily manipulated on probably any modern computer. Binning by 15-minute intervals reduces the data set to a nearly-trivial 3,575 observations, but the graphical summary becomes granular and less appealing at that resolution.

@@

# binnedRdm <- qplot(data=air.binned2.rdm, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
# ggsave("AirlineRdmBinning15mins.png", binnedRdm, width=8, height=8, dpi=600)
@

%--------------------------------------------------------------------------------------

\section{Conclusions and Future Work}

Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system. Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information.

We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient. The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid problematic visual features such as binning caused stripping. In setting binning parameters for each spatial dimension, binwidth should be an integer multiple of the data resolution and the origin should be offset half the data resolution below the lowest observed value.
